Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
finish java formal

Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\label{sec:isaformal}.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TNewStatic}
    {\TFieldStatic}

@@

    \BMethod{}
  \end{mathpar}
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
    \RFieldElimRefl{}
%
%    \RLet{}
%

    \RLetHint{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
  \label{main:figure:javatyping}
\end{figure}

In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
type hints to help the compiler eliminate reflective calls.
In this section, first we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this we add notation for
\emph{non-reflective} Java forms that carry sufficient tag information 
to unambiguously identify a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.

We model Java interoperability in a restricted setting without class inheritance,
overloading or Java Generics.

@@

We extend the syntax with Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form 
These come without typing or evaluation rules---Typed Clojure 
requires the Clojure compiler to statically resolve all Java interactions
in typed code, so they are implicitly type errors
(as demonstrated in section~\ref{sec:overviewjavainterop}).
We add type hinted syntax for local variables and let-bindings.

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
$$
{\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});
$$
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with signature
$$
{\classhint{2}}\ \fld{};
$$
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor
$$
{\classhint{}}\ ({\overrightarrow{\classhint{i}}});
$$

@@

$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{} and
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$

  $$
    \RFieldElimRefl{}
  $$

For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
rewrites the target expression and uses the result to infer a class type hint as the field target.
\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
field name and returns the actual target and result type hints which we store in the rewritten rule.

We then extract these type hints in T-FieldStatic for typing checking 
(figure~\ref{main:figure:javatyping}), which converts type hints
with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).

Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
we assume methods can return \nil{} and we assume constructors cannot.

We rewrite to our internal language before type checking and remaining reflective calls are type errors.

@@

{\convertjavatypefigure{figure}{\footnotesize}}

>>>>

Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\ref{sec:isaformal}.

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
 \classtablelookupsyntax{}
  \begin{mathpar}
    {\TNewStatic}
    {\TFieldStatic}

@@

    \BMethod{}
  \end{mathpar}
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
%    \RFieldElimRefl{}
%
%    \RLet{}
%

%    \RLetHint{}
  \end{mathpar}
  \convertjavatypegrammar{}
  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
  \label{main:figure:javatyping}
\end{figure}

In section~\ref{sec:overviewjavainterop} we discussed the role of type hints
to help eliminate reflective calls.
In this section, we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this, first we define notation for
\emph{non-reflective} Java forms that 
unambiguously call a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.
Finally we give typing rules that model how
Typed Clojure interacts with non-reflective calls.

We present Java interoperability in a restricted setting without class inheritance,
overloading or Java Generics.

@@

We extend the syntax with Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
These come without typing or evaluation rules---Typed Clojure 
demands statically resolved Java interactions
in typed code
(as demonstrated in section~\ref{sec:overviewjavainterop}).
We add type hinted syntax for local variables and let-bindings.

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

To obtain non-reflective calls we define our rewrite relation
$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

Type hint inference is modelled by
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$

To demonstrate we consider the expression 
$$
\fieldexp{getParent}{\hinted{\File}{\x{}}}.
$$
where 
$$
\ct = \{ \File \mapsto \{ \ctfieldsliteral \mapsto \{\ctfldentry{getMethod}{\String}\} \} \}
$$

R-FieldElimRefl eliminates a reflective field.
  $$
    \RFieldElimRefl{}
  $$
First the target is rewritten \rewrite {\taenv{}} {\hinted{\File}{\x{}}}{\hinted{\File}{\x{}}}
then we infer the type hint on the new expression \tajudgement {\taenv{}} {\hastype {\hinted{\File}{\x{}}} {\File}}
We call \fieldtypeliteral{} (figure~\ref{main:figure:javatyping})
to determine the exact signature of the field:
\fieldtype{\ct{}}{\File}{getParent}{\File}{\String}
and return the non-reflective field 
$$
\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}.
$$

Now we present the typing rules (figure~\ref{main:figure:javatyping}).
To briefly finish our example, let us assume \propenv{} = {\isprop{\File}{\x{}}},
then
T-FieldStatic converts the target Java type to a non-nilable
type \javatotc {\File} {\File}} and checks the target against it 
\judgement{\propenv{}}{\hastype{\x{}}{\File}}, and finally converts the return Java type 
to a nilable type and \javatotc {\String} {\Union{\Nil{}}{\String}}
to finish with 
$$
\judgementtwo{\propenv{}}{\hastype{\fieldstaticexp{\File}{\String}{getParent}{\hinted{\File}{\x{}}}}{\Union{\Nil{}}{\String}}}.
$$

The other rules T-MethodStatic and T-NewStatic work similarly, and vary
primarily in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method simply evaluate their
arguments and call the relevant JVM operation---section~\label{sec:metatheory}
states our exact assumptions about each one.

@@
